---
title: "Hidden Markov Models"
author: "First  Last Name"
date: "`r Sys.Date()`" #  
output:
  pdf_document:
      number_sections: true
spacing: double
header-includes:
- \usepackage{fancyhdr}
- \pagestyle{fancy}
- \fancyhead[L]{Title of Project}
- \fancyfoot[CO,CE]{}
- \fancyfoot[LE,RO]{\thepage}
- \usepackage{setspace}\doublespacing
---

\newpage 
\tableofcontents 
\newpage

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

\newpage

# Introduction

Hidden Markov Models (HMMs) are useful tools in stochastic modeling, especially when we study data that change over time but come from processes we cannot directly observe. Many real systems work this way. Examples include DNA sequences, financial trends, communication signals, and spoken language. In these situations, we only see the outcomes, while the actual states that generate them remain hidden. Because the observations may also contain noise, HMMs provide a structured and mathematical way to describe how such systems behave. With their combination of probability ideas and efficient algorithms, HMMs have become widely used for analyzing sequential data.

The idea behind an HMM comes from the Markov property, which says that the next state of a system depends only on its current state. Rabiner’s well-known 1989 paper helped define the modern framework by clearly separating the hidden Markov chain from the emission process that produces the observations. Thinking about HMMs in this way leads to three key questions:

1:How likely is a given sequence of observations? 2:Which hidden states most likely produced the sequence? 3:How can we learn the model parameters from data alone?

These questions correspond to the Forward–Backward algorithm, the Viterbi method, and the Baum–Welch procedure.

# Project Overview

In this project, we study HMMs from both theoretical and computational perspectives. We begin by defining the model structure and explaining the three main inference problems. After discussing the theory, we introduce the algorithms that solve these problems. We then use R to run simulations, estimate the model parameters, and recover the hidden states. By combining mathematical reasoning with practical computation, this project shows how HMMs can reveal hidden patterns in sequential data and why they are an important tool in stochastic process analysis.

# Basic Theory of Hidden Markov Chains

A Hidden Markov Model can be thought of as a regular Markov chain, except that the states are not directly visible. Instead, we only get to see some observations that are produced by these hidden states. To understand where the model comes from, it helps to first recall what a Markov chain is. A Markov chain is a sequence of random variables $X_1, X_2, \ldots , X_T$ that satisfies the **Markov property**:

$$P(X_{t+1}=j \mid X_t=i, X_{t-1},\ldots,X_1) = P(X_{t+1}=j \mid X_t=i).$$

This means that once we know the current state, anything further in the past does not matter for predicting the future states. The chain has a finite set of states,$$ S=\{1,2,\ldots,N\},$$ and a transition matrix $$ A = [a_{ij}], \qquad a_{ij} = P(X_{t+1}=j \mid X_t=i),$$ with each row's summation to one. This gives the full dynamics of a standard Markov chain.

What makes an HMM different is that we cannot observe $X_t$ directly. Instead, each hidden state produces an observable output. Let the set of all possible observations be $$ V=\{v_1, v_2, \ldots, v_M\}, $$ and let $$ b_j(k) = P(O_t = v_k \mid X_t = j)$$ represent the chance of seeing symbol $v_k$ when the chain is in hidden state $j$. The model also requires an initial distribution $$ \pi_i = P(X_1 = i),$$ with $$ \sum_{i=1}^N \pi_i = 1.$$ Together, the parameters of an HMM are written as $$ \lambda = (A, B, \pi). $$

Two assumptions are essential for the model to work properly:

1.  **State transitions follow the Markov property**,\
    meaning the hidden states themselves evolve like a first-order Markov chain: $$
    P(X_{t+1} \mid X_t, X_{t-1},\ldots) = P(X_{t+1} \mid X_t).
    $$

2.  **Observations depend only on the current hidden state**,\
    not on previous observations or states: $$ 
    P(O_t \mid X_1,\ldots,X_T, O_1,\ldots,O_{t-1} = P(O_t \mid X_t).
    $$

Under these assumptions, the joint probability of observing a particular hidden-state path $X = (x_1,\ldots,x_T)$ together with an observation sequence $O = (O_1,\ldots,O_T)$ can be written in a very structured way: $$
P(X,O \mid \lambda)
= \pi_{x_1} b_{x_1}(O_1)
  \prod_{t=2}^{T} a_{x_{t-1},x_t} \, b_{x_t}(O_t).
$$

This factorization is important because it avoids dealing with all $N^T$ possible state paths. It also explains why dynamic programming (like the Forward–Backward and Viterbi algorithms) works so well for HMMs.

HMMs are popular mainly because this setup gives enough structure to make inference possible, even though the true states are hidden. Many real-world systems—biological, financial, or communication-related—naturally fit into this framework, which is the reason why the model appears in so many applications.

# Three Fundamental Problems
When working with a Hidden Markov Model, there are three central questions that come up almost immediately. These questions describe what we want to learn from the observation sequence, even though the underlying states are hidden.
## 1.How likely is a given sequence of observations?
Suppose we observe a sequence
$$
O = (O_1, O_2, \ldots, O_T).
$$
The first thing we often want to know is simply: *How likely is it that our model produced this sequence?*  
This means we want to compute
$$
P(O \mid \lambda).
$$

In theory, we could sum over every possible hidden-state path:
$$
P(O \mid \lambda)
= \sum_{x_1,\ldots,x_T}
\pi_{x_1} b_{x_1}(O_1)
\prod_{t=2}^{T} a_{x_{t-1},x_t} \, b_{x_t}(O_t).
$$

But this involves $N^T$ possibilities, which is basically impossible to compute for any reasonable $T$.  
The **Forward algorithm** gives us a much faster way by using a recursive calculation instead of brute force.

## 2.Which hidden states most likely produced the sequence?

Another common question is: *Given what we observed, which hidden states were most likely responsible?*  
That is, we want a state sequence
$$
X^\* = \arg\max_X \, P(X \mid O, \lambda).
$$

Rather than checking every path, the **Viterbi algorithm** works through the sequence step by step and tracks the best partial path as it goes. By the end, it gives us the most likely full sequence.

## 3.How can we learn the model parameters from data alone?

If we only know $O$ but not the parameters $(A, B, \pi)$, then we want to estimate these in a way that makes the observed data as likely as possible. Formally,
$$
\lambda^\* = \arg\max_{\lambda} \, P(O \mid \lambda).
$$

This is hard to do directly since the states are hidden.  
The **Baum–Welch algorithm** (an EM method) tackles this by alternating between estimating hidden-state probabilities and updating the parameters until things stop changing much.

# Algorithms for Hidden Markov Models

## Notation and Model Setup

We consider a HMM with:

-   $N$ hidden states $X_t \in \{1, \dots, N\}$
-   $M$ possible observation symbols $o_t \in \{1, \dots, M\}$
-   Initial state distribution $\boldsymbol{\pi}=(\pi_1, \dots, \pi_N)$
-   Transition matrix $A=(a_{ij})$, where $$
    a_{ij}=P(X_t=j \mid X_{t-1}=i)
    $$
-   Emission matrix $B=(b_j(k))$, where $$
    b_j(k)=P(o_t=k \mid X_t=j)
    $$

We denote an observed sequence by $O=(o_1, \dots, o_T)$ and the corresponding hidden state sequence by $X=(X_1, \dots, X_T)$.

In the numerical examples below, we will use the following simple HMM (with $N=2$ states and $M=3$ observation symbols):

```{r hmm-params, echo=TRUE}
# Transition matrix A (rows = from-state, cols = to-state) 
A <- matrix(c(0.7, 0.3,
              0.4, 0.6),
            nrow = 2, byrow = TRUE)

# Emission matrix B(rows = state, cols = observation symbol)

B <- matrix(c(0.1, 0.4, 0.5, 
              0.6, 0.3, 0.1), 
            nrow = 2, byrow = TRUE)

# Initial distribution pi

pi <- c(0.6, 0.4)

# Example observation sequence (coded as 1, 2, 3)

obs <- c(1, 3, 2) 
obs
```

## 1. Forward Algorithm

### 1.1 Purpose (Evaluation Problem)

The forward algorithm computes the likelihood of an observation sequence\
$O = (o_1, o_2, \dots, o_T)$ under a given HMM $\lambda = (\boldsymbol{\pi}, A, B)$, that is $$
P(O \mid \lambda).
$$ This is known as the **evaluation problem** and it answers the question:

> How well does this HMM explain the observed data?

### 1.2 Intuition

Because the hidden states $X_t$ are not observed, many different hidden state sequences $X = (X_1, \dots, X_T)$ could have produced the same observation sequence $O$. A naive approach would sum the probability over all possible hidden-state sequences, but the number of such sequences grows exponentially with $T$.

The forward algorithm avoids this by using **dynamic programming**. At each time $t$, for each state $i$, we maintain a single quantity:

> The probability of being in state $i$ at time $t$, **and** having generated the observations up to time $t$.

This is the **forward variable**.

### 1.3 Mathematical Formulation

For each time $t = 1, \dots, T$ and state $i = 1, \dots, N$, define the forward variable $$
\alpha_t(i) = P(o_1, o_2, \dots, o_t,\; X_t = i \mid \lambda).
$$

We can compute $\alpha_t(i)$ recursively in three steps.

#### Initialization ($t = 1$)

At time $t = 1$, $$
\alpha_1(i) = \pi_i \, b_i(o_1), \quad i = 1, \dots, N,
$$ where $\pi_i = P(X_1 = i)$ is the initial state probability and $b_i(o_1) = P(o_1 \mid X_1 = i)$ is the emission probability.

#### Recursion ($t = 2, \dots, T$)

For time $t = 2, \dots, T$ and each state $i = 1, \dots, N$, $$
\alpha_t(i)
= \left( \sum_{j=1}^{N} \alpha_{t-1}(j)\, a_{j i} \right) b_i(o_t),
$$ where $a_{ji} = P(X_t = i \mid X_{t-1} = j)$ is the transition probability from state $j$ to state $i$.

#### Termination

After processing all $T$ observations, the likelihood of the entire sequence is $$
P(O \mid \lambda) = \sum_{i=1}^{N} \alpha_T(i).
$$

### 1.4 R Implementation

We now implement the forward algorithm in R as a function. This function assumes that the HMM parameters $A$, $B$, and $\boldsymbol{\pi}$, as well as the observation sequence `obs`, are already defined (as in the *Notation and Model Setup* section).

```{r forward-func, echo=TRUE}
forward <- function(A, B, pi, obs) {
  N <- length(pi) # number of states
  Tobs <- length(obs) # length of the observation sequence

  # alpha[t, i] = alpha_t(i) 
  alpha <- matrix(0, nrow = Tobs, ncol = N)

  # 1. Initialization
  alpha[1, ] <- pi * B[, obs[1]]

  # 2. Recursion
  for (t in 2:Tobs) {
    for (i in 1:N) { 
      alpha[t, i] <- sum(alpha[t - 1, ] * A[, i]) * B[i, obs[t]] 
    } 
  }

  # 3. Termination: likelihood
  likelihood <- sum(alpha[Tobs, ])
  list(alpha = alpha, likelihood = likelihood) 
}
```

### 1.5 Example: Evaluating the Toy HMM

We now apply the forward algorithm to the simple HMM defined in the *Notation and Model Setup* section. The following R code computes the forward variables $\alpha_t(i)$ and the likelihood $P(O \mid \lambda)$:

```{r forward-example, echo=TRUE}
res_forward <- forward(A, B, pi, obs)

# Forward variables alpha_t(i)
res_forward$alpha

# Likelihood P(O | lambda)
res_forward$likelihood
```

## 2. Backward Algorithm

### 2.1 Purpose (Support for Learning and Smoothing)

The backward algorithm is a dynamic programming procedure that propagates information **from the end of the sequence to the beginning**. It is mainly used to:

-   support **parameter learning** (Baum–Welch algorithm), and\
-   compute **posterior probabilities** of hidden states (via the forward–backward algorithm).

It provides an alternative way to compute the likelihood $P(O \mid
\lambda)$ that is mathematically equivalent to the forward algorithm.

### 2.2 Intuition

While the forward algorithm answers:

> “Given what I have seen so far, how probable is it to be in each state right now?”

the backward algorithm answers:

> “If I am in state $i$ at time $t$, how likely is it to generate **all the remaining future observations**?”

So:

-   **Forward** looks from left to right (past → present).\
-   **Backward** looks from right to left (future → present).

By combining these two views, we can later compute how likely it is to have been in each state at each time, given the entire observation sequence.

### 2.3 Mathematical Formulation

For each time $t = 1, \dots, T$ and state $i = 1, \dots, N$, define the backward variable $$
\beta_t(i) = P(o_{t+1}, o_{t+2}, \dots, o_T \mid X_t = i,\; \lambda),
$$ the probability of observing the **future** observations from time $t+1$ to $T$, given that the hidden state at time $t$ is $i$.

We compute $\beta_t(i)$ backwards in time.

#### Initialization ($t = T$)

At the final time $T$, there are no observations after $o_T$, so we set $$
\beta_T(i) = 1, \quad i = 1, \dots, N.
$$

This corresponds to the idea that “the probability of the empty sequence” is 1.

#### Recursion ($t = T-1, \dots, 1$)

For $t = T-1, T-2, \dots, 1$ and each state $i = 1, \dots, N$,

$$
\beta_t(i)
= \sum_{j=1}^{N} a_{i j} \, b_j(o_{t+1}) \, \beta_{t+1}(j),
$$

where:

-   $a_{i j} = P(X_{t+1} = j \mid X_t = i)$ is the transition probability from state $i$ to state $j$,
-   $b_j(o_{t+1}) = P(o_{t+1} \mid X_{t+1} = j)$ is the emission probability of the next observation from state $j$,
-   $\beta_{t+1}(j)$ is the backward variable at the next time step.

#### Likelihood from Backward Variables

At time $t = 1$, the overall likelihood of the observation sequence can be expressed in terms of $\beta_1(i)$ as

$$
P(O \mid \lambda)
= \sum_{i=1}^{N} \pi_i \, b_i(o_1) \, \beta_1(i).
$$

This value should be **identical** to the likelihood computed by the forward algorithm.

### 2.4 R Implementation

We now implement the backward algorithm in R. The function below returns both the matrix of backward variables $\beta_t(i)$ and the likelihood $P(O \mid \lambda)$ computed from the backward variables.

```{r backward-func, echo=TRUE}
backward <- function(A, B, pi, obs) {
  N    <- length(pi)        # number of states
  Tobs <- length(obs)       # length of the observation sequence
  
  # beta[t, i] = beta_t(i)
  beta <- matrix(0, nrow = Tobs, ncol = N)
  
  # 1. Initialization
  beta[Tobs, ] <- 1
  
  # 2. Recursion: t = T-1, ..., 1
  if (Tobs >= 2) {
    for (t in seq(Tobs - 1, 1, by = -1)) {
      for (i in 1:N) {
        beta[t, i] <- sum(A[i, ] * B[, obs[t + 1]] * beta[t + 1, ])
      }
    }
  }
  
  # 3. Likelihood using beta_1(i)
  likelihood <- sum(pi * B[, obs[1]] * beta[1, ])
  list(beta = beta, likelihood = likelihood)
}
```

### 2.5 Example: Evaluating the Toy HMM

We now apply the backward algorithm to the simple HMM. The following R code computes the backward variables $\beta_t(i)$ and the likelihood $P(O \mid \lambda)$:

```{r backward-example, echo=TRUE}
res_backward <- backward(A, B, pi, obs)

# Backward variables beta_t(i)
res_backward$beta

# Likelihood P(O | lambda)
res_backward$likelihood
```

## 3. Viterbi Algorithm

### 3.1 Purpose (Decoding Problem)

The Viterbi algorithm solves the **decoding problem** for hidden Markov models:

What is the single most likely sequence of hidden states $X = (X_1, X_2, \dots, X_T)$ that produced the observation sequence $O$?

Unlike the forward algorithm (which sums over all paths), the Viterbi algorithm finds **one best path**: the most probable hidden state trajectory.

### 3.2 Intuition

Imagine driving with a GPS:

-   There are many possible paths to your destination.
-   The forward algorithm considers **all** possible routes.
-   The Viterbi algorithm tries to find **the one best route**.

Viterbi works by keeping track of:

1.  The highest probability of any path ending in each state at each time. This is stored in a table called $$
    \delta_t(i).
    $$

2.  The “backpointer” that records which previous state produced that maximum.\
    This is stored in a table $$
    \psi_t(i).
    $$

At the end, we trace backward through $\psi_t(i)$ to recover the most likely state path.

### 3.3 Mathematical Formulation

Define the **Viterbi variable**:

$$
\delta_t(i)
= \max_{x_1,\dots,x_{t-1}}
P(X_1=x_1,\dots,X_t=i,\,o_1,\dots,o_t \mid \lambda),
$$

the highest probability of any path that ends in state $i$ at time $t$ and generates observations up to $t$.

We also define the **backpointer**:

$$
\psi_t(i)
= \arg\max_j \left[ \delta_{t-1}(j) a_{ji} \right].
$$

#### Initialization ($t = 1$)

$$
\delta_1(i) = \pi_i \, b_i(o_1),
\qquad
\psi_1(i) = 0.
$$

#### Recursion ($t = 2,\dots,T$)

$$
\delta_t(i)
=
\max_{j=1,\dots,N} [\delta_{t-1}(j) a_{ji}]
b_i(o_t),
$$

$$
\psi_t(i)
=
\arg\max_{j=1,\dots,N} \left[ \delta_{t-1}(j) a_{ji} \right].
$$

#### Termination

$$
P^* = \max_{i} \delta_T(i),
\qquad
X_T^* = \arg\max_{i} \delta_T(i).
$$

#### Path backtracking

For $t = T-1,\dots,1$:

$$
X_t^* = \psi_{t+1}(X_{t+1}^*).
$$

The resulting sequence $$
(X_1^*, X_2^*, \dots, X_T^*)
$$ is the most likely hidden-state path.

### 3.4 R Implementation

```{r viterbi-func, echo=TRUE}
viterbi <- function(A, B, pi, obs) {
  N    <- length(pi)
  Tobs <- length(obs)
  
  delta <- matrix(0, nrow = Tobs, ncol = N)
  psi   <- matrix(0, nrow = Tobs, ncol = N)
  
  # 1. Initialization
  delta[1, ] <- pi * B[, obs[1]]
  psi[1, ]   <- 0
  
  # 2. Recursion
  for (t in 2:Tobs) {
    for (i in 1:N) {
      probs <- delta[t - 1, ] * A[, i]
      psi[t, i]   <- which.max(probs)
      delta[t, i] <- max(probs) * B[i, obs[t]]
    }
  }
  
  # 3. Termination
  P_star <- max(delta[Tobs, ])
  X_star <- numeric(Tobs)
  X_star[Tobs] <- which.max(delta[Tobs, ])
  
  # 4. Backtracking
  for (t in (Tobs - 1):1) {
    X_star[t] <- psi[t + 1, X_star[t + 1]]
  }
  
  list(path = X_star, probability = P_star,
       delta = delta, psi = psi)
}
```

### 3.5 Example: Most likely State Sequence in Toy HMM

We now apply the Viterbi algorithm to the simple HMM.

```{r viterbi-example, echo=TRUE}
res_vit <- viterbi(A, B, pi, obs)

# Most likely hidden state sequence
res_vit$path

# Probability of this path
res_vit$probability

# Delta and psi tables
res_vit$delta
res_vit$psi
```

## 4. Baum-Welch Algorithm

# Computation for Hidden Markov Models

# Applications

# Conclusion

# References
Hughey, R., & Krogh, A. (1996). Hidden Markov models for sequence analysis: extension and analysis of the basic method. Bioinformatics, 12(2), 95-107.

Eddy, S. R. (2004). What is a hidden Markov model?. Nature biotechnology, 22(10), 1315-1316.

Schuster‐Böckler, B., & Bateman, A. (2007). An introduction to hidden Markov models. Current protocols in bioinformatics, 18(1), A-3A.
