---
title: "Hidden Markov Models"
author: "First  Last Name"
date: "`r Sys.Date()`" #  
output:
  pdf_document:
      number_sections: true
spacing: double
header-includes:
- \usepackage{fancyhdr}
- \pagestyle{fancy}
- \fancyhead[L]{Title of Project}
- \fancyfoot[CO,CE]{}
- \fancyfoot[LE,RO]{\thepage}
- \usepackage{setspace}\doublespacing
---

\newpage 
\tableofcontents 
\newpage

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

\newpage

# Introduction

What is a Hidden Markov Model? Why does it matter? What problem does it solve? Why did you choose this topic?

# Basic Theory of Hidden Markov Chains

# Three Fundamental Problems

## 1. Evaluation

## 2. Decoding

## 3. Learning

# Algorithms for Hidden Markov Models

## Notation and Model Setup

We consider a HMM with:

-   $N$ hidden states $X_t \in \{1, \dots, N\}$
-   $M$ possible observation symbols $o_t \in \{1, \dots, M\}$
-   Initial state distribution $\boldsymbol{\pi}=(\pi_1, \dots, \pi_N)$
-   Transition matrix $A=(a_{ij})$, where $$
    a_{ij}=P(X_t=j \mid X_{t-1}=i)
    $$
-   Emission matrix $B=(b_j(k))$, where $$
    b_j(k)=P(o_t=k \mid X_t=j)
    $$

We denote an observed sequence by $O=(o_1, \dots, o_T)$ and the corresponding hidden state sequence by $X=(X_1, \dots, X_T)$.

In the numerical examples below, we will use the following simple HMM (with $N=2$ states and $M=3$ observation symbols):

```{r hmm-params, echo=TRUE}
# Transition matrix A (rows = from-state, cols = to-state) 
A <- matrix(c(0.7, 0.3,
              0.4, 0.6),
            nrow = 2, byrow = TRUE)

# Emission matrix B(rows = state, cols = observation symbol)

B <- matrix(c(0.1, 0.4, 0.5, 
              0.6, 0.3, 0.1), 
            nrow = 2, byrow = TRUE)

# Initial distribution pi

pi <- c(0.6, 0.4)

# Example observation sequence (coded as 1, 2, 3)

obs <- c(1, 3, 2) 
obs
```

## 1. Forward Algorithm

### 1.1 Purpose (Evaluation Problem)

The forward algorithm computes the likelihood of an observation sequence\
$O = (o_1, o_2, \dots, o_T)$ under a given HMM $\lambda = (\boldsymbol{\pi}, A, B)$, that is $$
P(O \mid \lambda).
$$ This is known as the **evaluation problem** and it answers the question:

> How well does this HMM explain the observed data?

### 1.2 Intuition

Because the hidden states $X_t$ are not observed, many different hidden state sequences $X = (X_1, \dots, X_T)$ could have produced the same observation sequence $O$. A naive approach would sum the probability over all possible hidden-state sequences, but the number of such sequences grows exponentially with $T$.

The forward algorithm avoids this by using **dynamic programming**. At each time $t$, for each state $i$, we maintain a single quantity:

> The probability of being in state $i$ at time $t$, **and** having generated the observations up to time $t$.

This is the **forward variable**.

### 1.3 Mathematical Formulation

For each time $t = 1, \dots, T$ and state $i = 1, \dots, N$, define the forward variable $$
\alpha_t(i) = P(o_1, o_2, \dots, o_t,\; X_t = i \mid \lambda).
$$

We can compute $\alpha_t(i)$ recursively in three steps.

#### Initialization ($t = 1$)

At time $t = 1$, $$
\alpha_1(i) = \pi_i \, b_i(o_1), \quad i = 1, \dots, N,
$$ where $\pi_i = P(X_1 = i)$ is the initial state probability and $b_i(o_1) = P(o_1 \mid X_1 = i)$ is the emission probability.

#### Recursion ($t = 2, \dots, T$)

For time $t = 2, \dots, T$ and each state $i = 1, \dots, N$, $$
\alpha_t(i)
= \left( \sum_{j=1}^{N} \alpha_{t-1}(j)\, a_{j i} \right) b_i(o_t),
$$ where $a_{ji} = P(X_t = i \mid X_{t-1} = j)$ is the transition probability from state $j$ to state $i$.

#### Termination

After processing all $T$ observations, the likelihood of the entire sequence is $$
P(O \mid \lambda) = \sum_{i=1}^{N} \alpha_T(i).
$$

### 1.4 Pseudocode

We can summarize the forward algorithm as follows.

1.  **Initialization**:\
    For each state $i = 1, \dots, N$: $$
    \alpha_1(i) \leftarrow \pi_i \, b_i(o_1).
    $$

2.  **Recursion**:\
    For $t = 2, \dots, T$ and each state $i = 1, \dots, N$: $$
    \alpha_t(i) \leftarrow
    \left( \sum_{j=1}^{N} \alpha_{t-1}(j)\, a_{j i} \right) b_i(o_t).
    $$

3.  **Termination**:\
    Compute the likelihood $$
    P(O \mid \lambda) \leftarrow \sum_{i=1}^{N} \alpha_T(i).
    $$

### 1.5 R Implementation

We now implement the forward algorithm in R as a function. This function assumes that the HMM parameters $A$, $B$, and $\boldsymbol{\pi}$, as well as the observation sequence `obs`, are already defined (as in the *Notation and Model Setup* section).

```{r forward-func, echo=TRUE}
forward <- function(A, B, pi, obs) {
  N <- length(pi) # number of states
  Tobs <- length(obs) # length of the observation sequence

  # alpha[t, i] = alpha_t(i) 
  alpha <- matrix(0, nrow = Tobs, ncol = N)

  # 1. Initialization
  alpha[1, ] <- pi * B[, obs[1]]

  # 2. Recursion
  for (t in 2:Tobs) {
    for (i in 1:N) { 
      alpha[t, i] <- sum(alpha[t - 1, ] * A[, i]) * B[i, obs[t]] 
    } 
  }

  # 3. Termination: likelihood
  likelihood <- sum(alpha[Tobs, ])
  list(alpha = alpha, likelihood = likelihood) 
}
```

### 1.6 Example: Evaluating the Toy HMM

We now apply the forward algorithm to the simple HMM defined in the *Notation and Model Setup* section. The model has:

-   $N = 2$ hidden states\
-   $M = 3$ observation symbols\
-   Transition matrix $A$\
-   Emission matrix $B$\
-   Initial distribution $\boldsymbol{\pi}$

The observation sequence used is\
$$
O = (1,\, 3,\, 2).
$$

The following R code computes the forward variables $\alpha_t(i)$ and the likelihood $P(O \mid \lambda)$:

```{r forward-example, echo=TRUE}
res_forward <- forward(A, B, pi, obs)

# Forward variables alpha_t(i)
res_forward$alpha

# Likelihood P(O | lambda)
res_forward$likelihood
```

## 2. Backward Algorithm

### 2.1 Purpose (Support for Learning and Smoothing)

The backward algorithm is a dynamic programming procedure that propagates information **from the end of the sequence to the beginning**. It is mainly used to:

-   support **parameter learning** (Baum–Welch algorithm), and\
-   compute **posterior probabilities** of hidden states (via the forward–backward algorithm).

It provides an alternative way to compute the likelihood $P(O \mid
\lambda)$ that is mathematically equivalent to the forward algorithm.

### 2.2 Intuition

While the forward algorithm answers:

> “Given what I have seen so far, how probable is it to be in each state right now?”

the backward algorithm answers:

> “If I am in state $i$ at time $t$, how likely is it to generate **all the remaining future observations**?”

So:

-   **Forward** looks from left to right (past → present).\
-   **Backward** looks from right to left (future → present).

By combining these two views, we can later compute how likely it is to have been in each state at each time, given the entire observation sequence.

### 2.3 Mathematical Formulation

For each time $t = 1, \dots, T$ and state $i = 1, \dots, N$, define the backward variable
$$
\beta_t(i) = P(o_{t+1}, o_{t+2}, \dots, o_T \mid X_t = i,\; \lambda),
$$
the probability of observing the **future** observations from time $t+1$ to $T$, given that the hidden state at time $t$ is $i$.

We compute $\beta_t(i)$ backwards in time.

#### Initialization ($t = T$)

At the final time $T$, there are no observations after $o_T$, so we set 
$$
\beta_T(i) = 1, \quad i = 1, \dots, N.
$$

This corresponds to the idea that “the probability of the empty sequence” is 1.

#### Recursion ($t = T-1, \dots, 1$)

For $t = T-1, T-2, \dots, 1$ and each state $i = 1, \dots, N$,

$$
\beta_t(i)
= \sum_{j=1}^{N} a_{i j} \, b_j(o_{t+1}) \, \beta_{t+1}(j),
$$

where:

-   $a_{i j} = P(X_{t+1} = j \mid X_t = i)$ is the transition probability from state $i$ to state $j$,
-   $b_j(o_{t+1}) = P(o_{t+1} \mid X_{t+1} = j)$ is the emission probability of the next observation from state $j$,
-   $\beta_{t+1}(j)$ is the backward variable at the next time step.

#### Likelihood from Backward Variables

At time $t = 1$, the overall likelihood of the observation sequence can be expressed in terms of $\beta_1(i)$ as

$$
P(O \mid \lambda)
= \sum_{i=1}^{N} \pi_i \, b_i(o_1) \, \beta_1(i).
$$

This value should be **identical** to the likelihood computed by the forward algorithm.

### 2.4 Pseudocode

We can summarize the backward algorithm as follows.

1.  **Initialization**:\
    For each state $i = 1, \dots, N$: 
    $$
    \beta_T(i) \leftarrow 1.
    $$

2.  **Recursion** (for $t = T-1, \dots, 1$):\
    For each state $i = 1, \dots, N$: 
    $$
    \beta_t(i) \leftarrow
    \sum_{j=1}^{N} a_{i j} \, b_j(o_{t+1}) \, \beta_{t+1}(j).
    $$

3.  **Likelihood**:\
    Compute 
    $$
    P(O \mid \lambda) \leftarrow
    \sum_{i=1}^{N} \pi_i \, b_i(o_1) \, \beta_1(i).
    $$

### 2.5 R Implementation

We now implement the backward algorithm in R. The function below returns both the matrix of backward variables $\beta_t(i)$ and the likelihood $P(O \mid \lambda)$ computed from the backward variables.

```{r backward-func, echo=TRUE}
backward <- function(A, B, pi, obs) {
  N    <- length(pi)        # number of states
  Tobs <- length(obs)       # length of the observation sequence
  
  # beta[t, i] = beta_t(i)
  beta <- matrix(0, nrow = Tobs, ncol = N)
  
  # 1. Initialization
  beta[Tobs, ] <- 1
  
  # 2. Recursion: t = T-1, ..., 1
  if (Tobs >= 2) {
    for (t in seq(Tobs - 1, 1, by = -1)) {
      for (i in 1:N) {
        beta[t, i] <- sum(A[i, ] * B[, obs[t + 1]] * beta[t + 1, ])
      }
    }
  }
  
  # 3. Likelihood using beta_1(i)
  likelihood <- sum(pi * B[, obs[1]] * beta[1, ])
  list(beta = beta, likelihood = likelihood)
}
```
### 2.6 Example: Evaluating the Toy HMM

We now apply the backward algorithm to the simple HMM. The following R code computes the backward variables $\beta_t(i)$ and the likelihood $P(O \mid \lambda)$:

```{r backward-example, echo=TRUE}
res_backward <- backward(A, B, pi, obs)

# Backward variables beta_t(i)
res_backward$beta

# Likelihood P(O | lambda)
res_backward$likelihood
```

## 3. Forward-Backward Algorithm

## 4. Viterbi Algorithm

## 5. Baum-Welch Algorithm

# Computation for Hidden Markov Models

# Applications

# Conclusion

# References
